---
title: "Elements of Advanced R Markdown"
author: "Sabrina Nardin"
date: "2024-12-02"
output: html_document
---


## Inline code

Load libraries and data, notice we are displaying all R messages here. We could also suppress them using one of the code chunks options summarized below.
```{r}
library(tidyverse)
library(rcis)
```

Show Gun deaths by age:
```{r}
youth <- gun_deaths %>% filter(age <= 65)
```

Notice you can show your code output inline in your Rmd document like this: Data for this plot come from the `gun_deaths` dataset. We have data about `r nrow(gun_deaths)` individuals killed by guns. Only `r nrow(gun_deaths) - nrow(youth)` are older than 65.

The plot below (line 31) shows the distribution of individuals younger than 65. The peak of the distribution is around 22 years of age, with a second smaller peak around 50 years of age.


## Code chunks names

A named code chunk:

```{r youth-dist, cache = TRUE}
youth %>%
  ggplot(aes(age)) + 
  geom_histogram(binwidth = 1) +
  ggtitle("Individuals (0-65 years old) killed by guns") +
  xlab("Age") +
  ylab("Frequency")
```

Advantages of naming your code chunks:

1. More easily navigate to specific chunks (using the drop-down code navigator in the bottom-left of the script editor). NB: Two code chunks cannot have the same name or the document won't knit!
1. Graphics produced by the chunks will have meaningful names
1. Can be used with `cache = TRUE`

What is `cache = TRUE`? 
It is an R Markdown option that saves the output of a code chunk to a file in memory. On subsequent runs of the code, R checks if the code chunk has changed; if not, it reuses the saved output instead of re-running the code again (see Ch. 27 of "R for Data Science," 1st ed.). This is particularly useful for expensive computations or scraping projects, as it avoids re-running them every time you execute the code.


## Code chunks options

Most common code chunks options:

**`eval = FALSE`** code is not evaluated; code appears in the report, results do not appear in the report. Useful to show an example code in your report, or when code has an error you want to show. Default is `eval = TRUE` 

**`include = FALSE`** code is evaluated; code does not appear in the report, nor do results. Useful when you do not want to clutter your report with too much code. Default is `include = TRUE`

**`echo = FALSE`** code is evaluated; code does not appear in the report, results appear in the report. Useful to show your output to people that are not interested in the code that produced it. Default is `echo = TRUE`

**`error = TRUE`** code is evaluated and output appears in the report even if there is an error. Useful to knit with errors. Default is `error = FALSE`

**`message = FALSE`** or **`warning = FALSE`** prevents messages or warnings from appearing in the report. Default is `message = TRUE` or `warning = TRUE`

For more code chunks options see:
* https://r4ds.had.co.nz/r-markdown.html#chunk-options
* http://yihui.name/knitr/options/


## Global options

Use global options if you want to change the default chunk options for **all code chunks** in your file. 

This is how you set it up:

* create a new code chunk at the top of the document
* inside that code chunk, type `knitr::opts_chunk$set()` with your options. For example, if we want to hide all code by default, and add default width and height for all plots, w type:

```{r global-opts}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 7,
  fig.height = 5)
```


## YAML header specifications 

### Add a table of content

Add a table of content (toc) with depth 2. Pay attention at the indentation, it has to be exactly as shown below: 
```{r, eval = FALSE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document
    toc: true
    toc_depth: 2
---
```


### Knit in two (or more) formats at the same time

You first need to specify the formats in your YAML headers, then run the `rmarkdown::render()` function from your Console to generate both formats simultaneously. 

For example if you want to knit as html and pdf, with a table of content for each, type the following in your YAML header:
```{r, eval = FALSE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document:
    toc: true
  pdf_document:
    toc: true
---
```

To knit them both at the same time in this document (which is titled `more_Rmd`), type the following in your Console: `rmarkdown::render("more_Rmd.Rmd", output_format = "all")`. Check the `intro_Rmd` tutorial for more on this. 

To minimize repetitions, you can rewrite the code above as follows:
```{r, eval = FALSE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document:
  pdf_document:
    toc: true
---
```


### Add a date

Add date as a simple string
```{r, eval = FALSE}
title: "Gun deaths"
output: html_document
date: "2024-12-02"
```

Add the date automatically using `lubridate`
```{r, eval = FALSE}
---
title: "Gun deaths"
output: html_document
date: "`r lubridate::today()`"
---
```

Add the date automatically using `Sys.Date`, which returns the current day in the current time zone. The default uses the format year/month/day
```{r, eval = FALSE}
---
title: "Gun deaths"
output: html_document
date: "`r Sys.Date()`"
---
```

Add the date automatically using `Sys.time`, which returns an absolute date-time value which can be converted to various time zones. The format below says we want the date as: numeric day, full name of the month, and full (4-digit) year, but it can be changed. 
```{r, eval = FALSE}
---
title: "Gun deaths"
output: html_document
date: "`r format (Sys.time(), '%d, %B, %Y')`"
---
```

Quick reference on date formats in R: https://gist.github.com/zross/7fbbc034459aeff36627

More on dates (our book): https://r4ds.had.co.nz/dates-and-times.html



## Practice

In a team of two students, please complete the following:

1. Suppress messages when importing libraries. To observe the difference, knit with and without suppressing the messages. Click on 
"Session > Restart R and Clear Output" to reload the libraries.

2. Give a name to the first two, currently unnamed, code chunks of this file.

3. Use the code chunk named as `youth-dist` to test the following code chunk options: eval, include, echo. Start with the first one, set it as TRUE (e.g., `eval = TRUE`), knit the document, then set it as FALSE. Observe the differences in the rendered documents. Repeat for the other options.

4. Modify the YAML header of this document so it adds the date automatically (vs. the current simple string). Test a couple of methods illustrated in this tutorial and then pick one.

5. Modify the YAML header of this document so it generates an `html_document` with a table of contents with a depth of 2. Try to change the depth to 1, then to 3. Do they both work? What you need to modify so both works correctly?

6. Knit this document as html, md, and pdf at the same time. Add a table of contents to all of them.

7. Add the following two Global code options to this document: set default figure width and height, and enable caching. 

